\documentclass[a4paper]{report}
\usepackage{../template}
\begin{document}
\begin{defi}
  Let $R$ be a commutative ring.
  \begin{enumerate}[(a)]
    \item A category $\A$ is called $R$-linear if
          \begin{enumerate}[(i)]
            \item $\forall X, Y \in \A, \A(X,Y)$ is an $R$-module $(\A(X,Y), 0_{X,Y}, +_{X,Y}, \cd_{X,Y})$.
                  \item $\forall X, Y, Z \in \A$ the composition map \begin{align*}
                    \A(X,Y) \tm \A(Y,Z) & \to \A(X,Z) \\
                    (\ph, \psi) &\mps \psi \c \ph
                  \end{align*}
                  is $R$-bilinear. (in particular $r(\psi \c \ph) = (r \cd \psi) \c \ph = \psi \c(r \ph)$).
          \end{enumerate}
    \item A functor $F: \A \to \A'$ between $R$-linear categories is called $R$-linear if $\forall X, Y \in \A$ the map $F: \A(X,Y) \to \A'(FX,FY)$ is $R$-linear.
          \item For $\Z$-linear (preadditive) categories $\A$ and $\A'$, the full subcategory $\Add(\A,\A') \subseteq \Fun(\A,\A')$ of $\Z$-linear (additive) functors between them is again a $\Z$-linear category, i.e. there is a natural addition on natural transforations.
  \end{enumerate}
\end{defi}

\begin{exmps*}
\begin{enumerate}[(a)]
  \item $\RMod$ is $\Z$-linear and in fact ($R$ is commutative) $R$-linear.
  \item $\A$ is $R$-linear $\iff \A\op$ is also $R$-linear.
  \item If $\A$ and $\A'$ are $R$-linear, then $\A \tm \A'$ is $R$-linear.
        \item Let $S$ be a (not necessarily commutative) ring and $\underline S$ the category \[(\{*\}, S, \dom = \{*\}, \cod = \{*\}, \id_{*} = \id_{S}, \c = \cd_{S})\] associated to $S$, then $\underline S$ is $\Z$-linear $(\Hom_{\underline S}(*,*) = S)$.
\end{enumerate}
\end{exmps*}

\begin{lemm}
  If $\A$ is $\Z$-linear, then for $X \in \A$ the following are equivalent:
  \begin{enumerate}[(i)]
\item $X$ is initial.
    \item $X$ is terminal.
          \item $\A(X,X)=0$ (in particular $\id_{X} = 0$)
  \end{enumerate}
\begin{proof}
Exercise.
\end{proof}
\end{lemm}


\begin{defi}
  An $X \in \A$ satisfying conditions of Lemma 2 is called a \emph{zero object}
\end{defi}
\begin{nota*}
The zero object of $\A$ (if it exists) is unique up to unique isomorphism so we just write $0$ or $0_{A}$ for this object.
\end{nota*}

\begin{lemm}
  Let $\A$ be $\Z$-linear with a 0-object and $X, Y \in \A$, then the zero map $0_{X,Y} \in \A(X,Y)$ (remember $\A(X,Y)$ is an abelian group) is equal to the composition $X \xrightarrow{\ex !} 0\xrightarrow{\ex !} Y$
  \begin{proof}
Exercise.
  \end{proof}
\end{lemm}

\begin{prop}
  Let $\A$ be $\Z$-linear, then for $X_{1}, X_{2} \in \A$ the following are equivalent:
  \begin{enumerate}[(i)]
    \item The product $X_{1} \mathrel{\Pi} X_{2}$ exists.
    \item The coproduct $X_{1} \amalg X_{2}$ exists.
          \item $\ex Y \in \A, p_{1}, p_{2}: Y \to X_{i}$ and $\iota_{1}, \iota_{2}: X_{i} \to Y$ such that \[p_{i} \c \iota_{j} = \begin{cases}
            1_{X_{i}}, & i = j \\
            0, & i \ne j
          \end{cases}\]
  \end{enumerate}
\end{prop}
\begin{proof}
TODO.
\end{proof}
\begin{rem*}
In (iii) $(Y,\iota_{1}, \iota_{2})$ is the coproduct and $(Y, p_{1}, p_{2})$ is the product.
\end{rem*}

\begin{defi}
In a $\Z$-linear category we deonte $X \mathrel \Pi Y = X \amalg Y$ by $X \opl Y$ and call it the \emph{direct sum} of $X$ and $Y$.
\end{defi}

\begin{lemm}
  Let $\A$ and $\A'$ be $\Z$-linear and $F: \A \to \A'$ an additive functor, then
  \begin{enumerate}[(i)]
    \item If $0_{\A}$ exists then $0_{\A'}$ exists and $F(0_{\A}) \cong 0_{\A'}$.
          \item If $X, Y \in \A$ and $X \opl Y$ exists, then $FX \opl FY$ exists and $FX \opl FY \cong F(X\opl Y)$.
  \end{enumerate}
\begin{proof}
Exercise.
\end{proof}
\end{lemm}


\begin{defi}
A category $\A$ is called \emph{additive} (or $R$-linear additive) if $\A$ is $\Z$-linear (or $R$-linear), $0_{\A}$ exists and $\forall X, Y \in \A: X \opl Y$ exists.
\end{defi}

\begin{rem*}
If $\A$ is additive then the addition on $\A(X,Y)$ is determined by the composition map! (Morel II 1.2.4)
\end{rem*}

\section{Kernels and cokernels}
Recall that the equalizer of two morphisms $f, g: X \to Y$ is the limit of the diagram $X \overunderset fg \rightrightarrows Y$, so $Z \xrightarrow u X\overunderset fg \rightrightarrows Y$ such that for every $W \xrightarrow v X \overunderset fg \rightrightarrows Y$ with $f \c v = g \c v$, $v$ factors through $u$:
\[\begin{tikzcd}
Z \arrow[r, "u"]                     & X \arrow[r, "f", shift left] \arrow[r, "g"', shift right] & Y \\
W \arrow[ru, "v"'] \arrow[u, dashed] &                                                           &
\end{tikzcd}\]similarly the coequalizer is the colimit of $X\overunderset fg \rightrightarrows Y$.
\newline
In the category of $R$-modules, $\eq(f,g) = \ker(f - g)$ and $\coeq(f,g) = \cok(f - g)$. In particular $\ker f = \eq(f,0)$.

\begin{defi}
  Let $\A$ be a $\Z$-linear category and $f \in \A(X,Y)$.
  \begin{enumerate}[(a)]
    \item define $\ker f := \eq(f,0)$ and $\cok f = \coeq(f,0)$ (if they exist)
    \item If $\ker f$ exists, define the coimage as $X \to \coim f = \cok(\ker f \to X)$ (it might not exist).
          \item If $\coker f$ exists, define the image as $\im f \to Y = \ker(Y \to \cok f)$.
  \end{enumerate}
\end{defi}

\begin{exmp}
Let $X_{1}, X_{2} \in \A$ and $(X_{1} \opl X_{2}, \iota_{1}, \iota_{2}, p_{1}, p_{2})$ the direct sum. Then $\iota_{1} = \ker p_{2}, \iota_{2} = \ker p_{1}, p_{1} = \cok \iota_{2}, p_{2} = \cok \iota_{1}$
\end{exmp}

\begin{lemm}
  Kernels are monomorphisms and cokernels are epimorphisms.
\end{lemm}
\begin{proof}
  We only prove the statement for kernels. Let $\ker f \xrightarrow \iota X \overunderset f0 \rightrightarrows Y$, now assume $\iota \c \ph = \iota \c \ph'$
  \[\begin{tikzcd}
Z \arrow[r, "\varphi", shift left] \arrow[r, "\varphi'"', shift right] & \ker f \arrow[r, "\iota"] & X \arrow[r, "f", shift left] \arrow[r, "g"', shift right] & Y
\end{tikzcd}\]
By the definition of equalizer \begin{align*}
  \A(Z, \ker f) &\simeq \{\psi: Z \to X \mid f \c \psi = 0\} \\
  \ph &\mps \iota \c \ph
\end{align*}
is bijective, so $\ph = \ph'$.
\end{proof}


\begin{thm}Let $f \in \A(X,Y)$ and assume $\ker f, \cok f, \im f, \coim f$ exist. Then there exists a unique morphism $u: \coim f \to \im f$ such that $f$ is equal to the composition
  \[\begin{tikzcd}
      X \arrow[r, "c"] & \coim f \arrow[r, "\ex!u", dashed] & \im f \arrow[r, "d"] & Y
    \end{tikzcd}\]
  This is called the \emph{canonical factorization} of $f$ (or epi-mono factorization).
\end{thm}
\begin{rem*}
Note that by lemma 11, $c$ is an epimorphism and $d$ is a monomorphism.
\end{rem*}
\begin{proof}TODO.

\end{proof}
Note that in the category of $R$-modules, $u$ is an isomorphism.


\begin{defi}
  A category $\A$ is called \emph{abelian} if
  \begin{enumerate}[(i)]
    \item $\A$ is additive ($\Z$-linear, 0 exists and $X \opl Y$ exists)
    \item $\forall f \in \A(X,Y): \ker f, \coker f$ exist.
          \item $\forall f \in \A(X,Y)$ with canonical factorization $f = d \c u \c c$, $u$ is an isomorphism.
  \end{enumerate}
\end{defi}

\begin{exmp}
\begin{itemize}
  \item For any ring $R$ the category $\RMod$ is abelian.
        \item For rings $R, R'$ the category $\RMod_{R'}$ is abelian $( \cong {}_{R \otm_{\Z}(R')\op}\Mod )$.
\end{itemize}
\end{exmp}

\begin{exmp}
    \item Let $\A \subseteq \Ab$ be the fullsubcategory of finitely generated free $\Z$-modules, then $\A$ is additive.
    \item For $f: X \to Y$ in $\A$ the usual $\ker_{\Ab} f$ as abelian group is again a finitely generated free abelian group and is also the kernel in $\A$.
    \item The cokernel however might have torsion. In fact \[\coker_{\A} f = \fak{\coker_{\Ab} f}{\Tor(\coker_{\Ab} f)}\]
          So kernels and cokernels exists in $\A$, now let $f: \Z \to \Z, n \mps 2n$, the canonical factorization is \[\begin{tikzcd}
\Z \arrow[r, "\id"] & \Z \arrow[r, "u"]    & \Z \arrow[r, "\id"] & \Z \\[-20]
                    & n \arrow[r, maps to] & 2n                  &
\end{tikzcd}\]since $u$ is not an isomorphism, $\A$ is not abelian.
\end{exmp}


\begin{prop}
  Let $\C$ be a category, then
 \begin{enumerate}[(i)]
   \item $\C$ is $\Z$-linear $\iff \C\op$ is $\Z$-linear.
   \item $\C$ is additive $\iff \C\op$ is additive.
         \item $\C$ is abelian $\iff \C\op$ is abelian.
 \end{enumerate}
 Moreover (if they exist) $\ker_{\C} f = \coker _{\C\op} f$ and $\im_{\C} f = \coim_{\C\op} f$.
\end{prop}


\section{Abelian categories}
From now on let $\A$ be an abelian category.
\begin{rem*}
Let $f: X \to Y$ and $X \twoheadrightarrow \coim f \overunderset \cong u \rightarrow \im f \hookrightarrow Y$ be the canonical factorization. Then either $X \twoheadrightarrow \coim f \hookrightarrow Y$ or $X \twoheadrightarrow \im f \hookrightarrow Y$ (or anything else isomorphic to them) is called ``the'' canonical factorization for $f$.
\end{rem*}

\begin{prop}
  Let $X, Y \in \A$
  \begin{enumerate}[(a)]
    \item $\coker (0 \to X) = X \xrightarrow \id X$ and $\ker(X \to 0) = X \xrightarrow \id X$.
    \item $f: X \to Y$ is a monomorphism $\iff \ker f = 0 \iff$ the canonical factorization of $f$ is $X \xrightarrow \id X \overset f \hookrightarrow Y \iff f$ is a kernel.
    \item $g: C \to Y$ is an epimorphism $\iff \coker g= 0 \iff$ the canonical factorization of $g$ is $X \xrightarrow g Y \xrightarrow \id Y \iff g$ is a cokernel.
          \item $u$ is an isomorphism $\iff u$ is a monomorphism and an epimorphism.
  \end{enumerate}
\end{prop}


\begin{cor}
$X \xrightarrow a Z \xrightarrow b Y$ is the canonical factorization for $f = b \c a \iff a$  ia a monomorphism and $b$ an epimorphism.
\end{cor}

\begin{defi}
In an abelian category $f \in \A(X,Y)$ is called injective if $\ker f = 0$ and surjective if $\coker f = 0$.
\end{defi}

\begin{cor}
  Let $f \in \A(X,Y)$, then
 \begin{enumerate}[(a)]
\item $f$ is injective $\iff f$ is a monomorphism.
\item $f$ is surjective $\iff f$ is a epimorphism.
\item $f$ is injective and surjective $\iff f$ is a isomorphism.
 \end{enumerate}
\end{cor}


\section{Exactness}
\begin{lemm}
  Let $(g,f)$ be a composable pair of morphisms in $\A$ such that $g \c f = 0$, then $\ex$ a canonical injection $\im f \hookrightarrow \ker g$.
\end{lemm}
\begin{proof}
  Let $f: X \to Y$ and $g: Y \to Z$ and write the epi-mono factorization of $f$:
  \[\begin{tikzcd}
X \arrow[r, "c", two heads] & \coim f \arrow[r, "\cong"] & \im f \arrow[r, "d", hook] & Y \arrow[r, "g"] & Z
\end{tikzcd}\]and call the isomorphism in the middle $u: \coim f \cong \im f$.
Now $g \c d \c u \c c = 0 \imp g \c d \c u = 0$ since $c$ is an epimorphism and $u$ isomorphism $\imp g \c d = 0$. By the definition of the kernel $d$ must factor through $\ker g$:
\[\begin{tikzcd}
\im f \arrow[r, "e"] & \ker g \arrow[r, "\iota", hook] & Y \arrow[r, "g"] & Z
\end{tikzcd}\]
Since $d$ is injective: if $e \c \ph = c \c \ph ' \imp \iota \c c \c \ph = \iota \c c \c \ph' \imp d \c \ph = d \c \ph' \imp \ph = \ph ' \imp e$ is a monomorphism $\imp \im f \hookrightarrow \ker g$.
\end{proof}

\begin{defi}
\begin{enumerate}[(a)]
  \item A \emph{complex} in $\A$ is a family of morphisms $(d_{i}: X_{i} \to X_{i+1})_{i \in J}$ and $\emptyset \ne J \subseteq \Z$ an interval such that $\forall i \in J^{-}: d_{i+1} \c d_{i} = 0$ ($J^{-}:= J \setminus \sup J$), in other notation:
        \[ \cdots \to X_{i-1} \xto {d_{i-1}} X_{i} \xto {d_{i}} X_{i+1} \to \cdots\]such that $d_{i} \c d_{i-1} = 0$.
        \item An exact sequence in $\A$ (or an acyclic complex) is a complex such that $\forall i \in J^{-}$ the canonical monomorphism from lemma 21 $\im d_{i} \hookrightarrow \ker d_{i+1}$ is an isomorphism
\end{enumerate}
\end{defi}


\begin{prop}
  For any $f \in \A(X,Y)$ the sequence \[0 \to \ker f \xto \iota X \xto f Y \xto \pi \coker f \to 0\]
  is exact, where $\iota$ is the kernel-morphism and $\pi$ is the cokernel-morphism.
  \begin{proof}
\begin{itemize}
  \item Exactness at $\ker f$: Kernels are monomorphisms and so \[\ker \iota = 0 = \im(0 \to \ker f)\]
  \item Exactness at $X$:
        \[\im \iota \underset {16} = \ker f \xto \iota X = \ker f\]
        for $Y$ and $\coker f$ pass to $\A\op$.
\end{itemize}
  \end{proof}
\end{prop}


\begin{lemm}[Decomposing and concatenating exact sequences]\item
\begin{enumerate}[(a)]
        \item If
        \[
  \cdots \to A \xto a B \to 0 \quad \text{and} \quad
        0 \to B \xto b C \xto c D \to \cdots
        \]
        are exact at $B$ and $C$, then
        \[\cdots \to A \tto{b \c a} C \tto c D \to \cdots\]
        is exact at $C$, i.e. $c \c b \c a = 0$ and $\ker c = \im(b \c a)$.
  \item If
        \[
  \cdots \to A \xto a B \tto b C \to 0 \quad \text{and} \quad
        0 \to  C \xto c D \to \cdots
        \]
        are exact at $B$ and $C$, then
        \[\cdots \to A \tto a B \tto {c \c b} D \to \cdots\]
        is exact at $B$.

  \item If
        \[\cdots \to X \tto f Y \tto g Z \to \cdots\]is exact at $Y$, then:
        \begin{enumerate}[(i)]
          \item $\cdots \to X \to \im f \to 0$ and $0 \to \ker g \to Y \tto g Z \to \cdots$ are exact at $\im f, \ker g, Y.$
                \item $\cdots X \tto f Y \to \coker f \cong \coim g \to 0$ and $0 \to \im g \to Z \to \cdots$ are exact at $Y, \coker f, \im g$.
        \end{enumerate}
\end{enumerate}
\end{lemm}

\begin{exmp*}
  Prop 23 + Lemma 24 gives for $f \in \A(X,Y)$:
  \[0 \to \ker f \to X \to \coim f \to 0\]
  and
  \[0 \to \im f \to Y \to \coker f \to 0\]
  are exact sequences.
\end{exmp*}
\begin{defi*}
  An exact sequence of the form \[0 \to A \to B \to C \to 0\]
  is also called a \emph{short exact sequence} (s.e.s.)
\end{defi*}

\begin{lemm}
  Let $0 \to A \tto a B \tto b C \to 0$ be a complex in $\A$, then:
  \begin{enumerate}[(a)]
    \item The sequence
          \[\begin{tikzcd}
A \arrow[r, "a"] & B \arrow[r, "b"] & C \arrow[r] \arrow[rd] & 0 \\
                 &                  &                        & X
               \end{tikzcd}\]
          is (right) exact $\iff \forall X \in \A$
          \[0 \to \Hom_{\A}(C,X) \tto {b^{*} = - \c b} \Hom_{\A}(B,X) \tto {a^{*}} \Hom_{\A}(A,X)\]
          is exact in $\Ab$.
    \item The sequence
          \[0 \to A \tto a B \tto b C\]
          is (left) exact $\iff \forall X \in \A:$
          \[0 \to \Hom_{\A}(X,A) \tto{a_{*}} \Hom_{\A}(X,B) \tto {b_*} \Hom_{\A}(X,C)\]
          is exact in $\Ab$.
  \end{enumerate}
\begin{proof}
Stacks Project.
\end{proof}
\end{lemm}

\begin{prop}
  Let the following be a short exact sequence in $\A$
\[\begin{tikzcd}
0 \arrow[r] & A \arrow[r, "a"] & B \arrow[r, "b"] \arrow[l, "s", dashed, bend left] & C \arrow[r] \arrow[l, "t", dashed, bend left] & 0
\end{tikzcd}\]
Then the following are equivalent:
\begin{enumerate}[(i)]
\item $\ex s \in \A(B,A) : s \c a = 1_{A}$
  \item $\ex t \in \A(C,B) : b \c t = 1_{C}$
  \item $\ex s \in \A(B,A) \ex t \in \A(C, B) : s \c a = 1_{A}, b \c t = 1_{C}$ and $1_{B} = a \c s + t \c b$
  \item $B \cong A \opl C$ with respect to the morphisms from (iii).
\end{enumerate}

        \begin{proof}
          Exercise. (See 1.28)
        \end{proof}
\end{prop}

\begin{prop}
  Let $\A$ be an abelian category, then $\A$ possesses (all) finite limits and colimits, (finite meaning over finite index categories).
\begin{proof}
Exercise (See 2.46)
\end{proof}
\end{prop}

\begin{prop}[Special case of 27]
  If $\A$ is an abelian category, then $\A$ has pullbacks and pushouts. They are given by
  \[\mathrm{p.b.}\l( \begin{tikzcd}
                  & Y \arrow[d, "g"] \\
X \arrow[r, "f"'] & Z
\end{tikzcd}\r) = \ker (X \opl Y \tto{f - g} Z)\]
and
\[\mathrm{p.o.}\l(  \begin{tikzcd}
X \arrow[d, "g"'] \arrow[r, "f"] & Y \\
Z                                &
\end{tikzcd} \r) = \coker(X \tto {(f,-g)} Y \opl Z)\]
\begin{proof}
Exercise.
\end{proof}
\end{prop}

\begin{defi}
A commutative diagram \[\begin{tikzcd}
A \arrow[r, "a"] \arrow[d, "b"'] & B \arrow[d, "c"] \\
C \arrow[r, "d"']                & D
\end{tikzcd}\] in $\A$ is called
\begin{enumerate}[(a)]
  \item \emph{cartesian} if
        \[\begin{tikzcd}
            A \arrow[d, "b"'] \arrow[r, "a"] & B \\
            C
          \end{tikzcd} \cong \mathrm{p.b.} \l( \begin{tikzcd}
            & B \arrow[d, "c"] \\
            C \arrow[r, "d"'] & D
          \end{tikzcd} \r)\]
  \item \emph{cocartesian} if
        \[\begin{tikzcd}
            & B \arrow[d, "c"] \\
            C \arrow[r, "d"'] & D
          \end{tikzcd} \cong \mathrm{p.o.} \l( \begin{tikzcd}
            A \arrow[d, "b"'] \arrow[r, "a"] & B \\
            C
          \end{tikzcd} \r) \]
\end{enumerate}
\end{defi}


\begin{lemm}
  Let $(*)$ be a commutative diagram in $\A$
  \[\begin{tikzcd}
      A \arrow[r, "a"] \arrow[d, "b"'] & B \arrow[d, "c"] \\
      C \arrow[r, "d"']& D
    \end{tikzcd} \quad \quad (*)\]
  Then:
  \begin{enumerate}[(a)]
    \item $(*)$ is cartesian $\iff 0 \to A \tto{(a,b)} B \opl C \tto{c + d} D$ is exact.
    \item $(*)$ cocartesian $\iff A \to B \opl C \to D \to 0$ is exact.
    \item If $(*)$ is cartesian, then $\ker a \cong \ker d$ via $b$.
    \item If $(*)$ is cocartesian, then $\coker a \cong \coker d$ via $c$.
          \item If $(*)$ is cartesian and $d$ is surjective, then $(*)$ is cocartesian.
          \item If $(*)$ is cocartesian and $a$ is injective, then $(*)$ is cartesian.
  \end{enumerate}

\end{lemm}
\section{The homomorphism theorem and the isomorphism theorems}
\begin{defi}
\begin{enumerate}[(a)]
  \item On the set of monomorphisms to $X$ define an equivalence relation $(f: U \hookrightarrow X) \sim (g: V \hookrightarrow X) \iff \ex$ commutative diagram
        \[\begin{tikzcd}
U \arrow[rd, "f"', hook] \arrow[rr, "\alpha", dashed] &   & V \arrow[ld, "g", hook] \\
                                                      & X &
\end{tikzcd}\]
with $\a$ an isomorphism. Call the equivalence classes \emph{subobjects} of $X$.
  \item On the set of epimorphisms from $Y$ define an equivalence relation $(f: Y \twoheadrightarrow W) \sim (g: Y \twoheadrightarrow Z) \iff \ex$ commutative diagram
\[\begin{tikzcd}
                              & Y \arrow[ld, two heads, "f"'] \arrow[rd, two heads, "g"] &   \\
W \arrow[rr, "\beta"', dashed] &                                    & Z
\end{tikzcd}\]

        with $\b$ an isomorphism. Call the equivalence classes \emph{quotients} of $Y$.
\end{enumerate}
Also: Call a monomorphism $f: U \to X$ (epimorphism $g: Y \to W$) a subobject (quotient) meaning its equivalence class.
\end{defi}

\begin{prop}
  On subobjects $[U \tto f X], [V \tto X]$ of $X \in \A$ define a relation $[U \tto f X]\le [V \tto X] \iff \ex$ commutative diagram:

        \[\begin{tikzcd}
U \arrow[rd, "f"', hook] \arrow[rr, "\alpha", dashed, hook] &   & V \arrow[ld, "g", hook] \\
                                                      & X &
                                                    \end{tikzcd}\]with $\a$ a monomorphism. Then
\begin{enumerate}[(a)]
  \item $\le$ is a partial ordering on the class of subobjects.
  \item The infimum of $[U \tto f X]$ and $[V \tto g X]$ exists and is represented by the pullback \[\mathrm{p.b.}\l( \begin{tikzcd}
      & V \arrow[d] \\
      U \arrow[r] & X
                                                    \end{tikzcd} \r)\]
        together with the diagonal map $\mathrm{p.b.}(\cdots) \to X$. Write $U \cap V$.
  \item The supremum of $[U \tto f X]$ and $[V \tto g X]$ exists and is represented by the pushout
        \[\mathrm{p.o.}\l( \begin{tikzcd}
            U \cap V \arrow[d] \arrow[r]&  U\\
            V
          \end{tikzcd} \r)\]
        together with the canonical  arrow $U \cap V \to \mathrm{p.o.}(\cdots)$. Write $U + V$
\end{enumerate}
\end{prop}

\begin{rem}
  Now we can formulate the homomorphism and isomorphism theorems:
  \begin{enumerate}[(a)]
    \item \textbf{Homomorphism theorem:} Given $f \in \A(X,Y) \ex$ canonical factorization:
          \[\begin{tikzcd}
              X \arrow[r, "f"] \arrow[d, "\mathrm{can.}"'] & Y \\
              \displaystyle \fak X{\ker f} := \coim f \arrow[r, "\cong"'] & \im f \arrow[u, "\mathrm{can.}"']
            \end{tikzcd}\]
    \item \textbf{First isomorphism theorem:} For subobjects $U, V \le X$ we have \[\fak{U+V}U \cong \fak V{V \cap U}\]
    \item \textbf{Second isomorphism theorem:} Given a monomorphism $\iota : U \to X, \ex$ bijection:
          \[\l\{ V \mid U \le V \le X \r\} \leftrightarrow \l\{V \le \fak XU = \coim \iota\r\}\]
          and for $U, V$ subobjects with $\b : U \to V$ and $\iota: V/U \hookrightarrow X/U$, we have an isomorphism \[\fak XV \cong \fak{(X/U)}{(V/U)} = \coim \iota\]
  \end{enumerate}

\end{rem}


\begin{thm}[Snake lemma]
  TODO.

\end{thm}


\end{document}
